---
title: "Intro to R - Data Structures"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r setup, include=FALSE}
random_norms <- rnorm(100)
some_letters <- letters[1:10]
```

# R Primitives

## Quick Tour of Things You Need to Know
### Data Structures

> "Bad programmers worry about the code. 
> Good programmers worry about data structures and their relationships."
> - Linus Torvalds

* R's data structures can be described by their dimensionality, and their type.


|    | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

### Data Types

* Atomic vectors come in one of four types
* `logical` (boolean). Values: `TRUE` | `FALSE`
* `integer`
* `double` (often called numeric)
* `character`
* Rare types:
* `complex` 
* `raw`

## Manipulating Data Structures

### Subsetting Operators

* To create a vector, use `c`: `c(1, 4, 1, 3)`
* To create a list, use `list`: `list(1, 'hi', data.frame(1:10, letters[1:10]))`
* To subset a vector or list, use the `[ ]`
  - inside the brackets: 
    + positive integer vectors for indices you want 
    + negative integer vectors for indices you want to drop
    + logical vectors for indices you want to keep/drop (TRUE/FALSE)
    + character vectors for named vectors corresponding to which named elements you want to keep
    + subsetting a list with a single square bracket always returns a list
+ To subset a list and get back just that element, use `[[ ]]`


### Example

+ Use `c` to create a vector of numbers, separate them by using a comma.
+ You can also use the `seq` function to create a sequence of numbers. `seq` has three arguments: 1. start, 2. stop, 3. increments
+ For a sequence of numbers, you can use `:`
+ Use the single square brackets to subset the first element
+ Use the single square brackets to subset the second and fourth elements

```{r datastructs, exercise=TRUE}

vector <- c(1)

seq(0, 10, 2)

a <- 2
b <- 4
vector_num_1 <- a:b


```


### Object Representation

+ To find the type of an object, use `class` (_higher level representation_)
+ To find how the object is stored in memory, use `typeof` (_lower level representation_)

## Vectors 

Use the `typeof` function to explore the data structure's low-level structure, and the `class` object to see the object's higher level structure.

```{r vectors, exercise=TRUE}
random_norms <- rnorm(100)
typeof(random_norms)
class(random_norms)

some_letters <- letters[1:10]
typeof(some_letters)
class(some_letters)

int_vector <- c(1L, 2L, 3L)
typeof(int_vector)
class(int_vector)

booleans <- int_vector == 1
typeof(booleans)
class(booleans)
```

+ can you mix types in a vector?

+ what happens when you mix the numeric with letters?

```{r coercion, exercise=TRUE}
random_norms <- rnorm(100)
some_letters <- letters[1:10]

combine_char_num <- c(random_norms, some_letters)

# check the type of combine_char_num

```

## Lists

to combine types, make list

```{r list, exercise=TRUE}
random_norms <- rnorm(100)
some_letters <- letters[1:10]

list_char_num <- list(nums = random_norms, chars = some_letters)
typeof(list_char_num)
lapply(list_char_num, typeof)
```

## Matrices

matrices, same length, same type

```{r matrices, exercise=TRUE}
matrix_num <- matrix(rnorm(10), nrow = 5, ncol = 2)
# what happens if you mix types?
matrix_mix <- matrix(c(rnorm(10), letters[1:10]), nrow = 10, ncol = 2)
```

be careful about specifying matrices, as they will auto-fill when there is an insufficient amount of elements;

```{r matrices-filling}
matrix_num_reuse <- matrix(rnorm(11), nrow = 6, ncol = 4)
```


## data.frames

To mix elements in a rectangular object/table, use `data.frames`:

```{r dfs, exercise=TRUE}
df_char_num <- data.frame(chars = letters[1:10], nums = rnorm(10))
lapply(df_char_num, typeof)

## data.frames must have same length in each column

# df_char_num <- data.frame(chars = letters[1:8], nums = rnorm(10))

```

## Helpful Functions


```{r help, exercise=TRUE}

# see your workspace
ls()


# check working directory, change directory
getwd()
setwd(getwd())

# create a sequence of numbers
1:10
seq(1, 10)
seq(1, 10, 2)

# get help
?seq
help(seq)

# type tests

is.character(some_letters)
is.numeric(some_letters)
is.atomic(some_letters)
is.atomic(df_char_num)

# remove object from workspace
rm(some_letters)

# remove all visible objects

rm(list = ls())

```

