---
title: "Hands on - `dplyr` Methods"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---


# Data Manipulation with the `dplyr` Package

## Overview

Rather than describing the nitty gritty details of writing R code, I'd like you to get started at immediately writing R code.

As most of you are data scientists/data enthusiasts, I will showcase one of the most useful data manipulation packages in R, `dplyr`.
At the end of this session, you will have learned:

* How to manipulate data quickly with `dplyr` using a very intuitive _"grammar"_
* How to use `dplyr` to perform common exploratory analysis data manipulation procedures
* How to apply your own custom functions to group manipulations `dplyr` with `mutate()`, `summarise()` and `do()`
* Connect to remote databases to work with larger than memory datasets

## Why use dplyr? 
### The Grammar of Data Manipulation

* `dplyr` is currently the [most downloaded package](https://www.rdocumentation.org/packages/dplyr/versions/0.5.0?) from CRAN
* `dplyr` makes data manipulation easier by providing a few functions for the most common tasks and procedures
* `dplyr` achieves remarkable speed-up gains by using a C++ backend
* `dplyr` has multiple backends for working with data stored in various sources: SQLite, MySQL, bigquery, SQL Server, and many more
* `dplyr` was inspired to give data manipulation a simple, cohesive grammar (similar philosophy to `ggplot` - grammar of graphics)
* `dplyr` has inspired many new packages, which now adopt it's easy to understand syntax. 
* The recent packages `dplyrXdf` and `SparkR/sparklyr` brings much of the same functionality of `dplyr` to `XDF`s data and Spark `DataFrames`


## Tidy Data and Happier Coding
### Premature Optimization 

+ The most important parameter to optimize in a data science development cycle is YOUR time
+ It is therefore important to be able to write efficient code, quickly
+ Goals: writing fast code that is: portable, platform invariant, easy to understand, and easy to debug
    - __Be serious about CReUse__!

## Manipulation verbs

`filter`

:    select rows based on matching criteria

`slice`

:    select rows by number

`select`

:    select columns by column names

`arrange`

:    reorder rows by column values

`mutate`

:    add new variables based on transformations of existing variables

`transmute`

:    transform and drop other variables



## Aggregation verbs

`group_by`

:    identify grouping variables for calculating groupwise summary statistics


`count`

:    count the number of records per group


`summarise` | `summarize`

:    calculate one or more summary functions per group, returning one row of results per group (or one for the entire dataset)

## NYC Taxi Data
### Data for Class

* The data we will be examining in this module is dervided from the [NYC Taxi and Limousine Commission](http://www.nyc.gov/html/tlc/html/home/home.shtml)
* Data contains taxi trips in NYC, and includes spatial features (pickup and dropoff neighborhoods), temporal features, and monetary features (fare and tip amounts)
* The dataset for this module is saved as an _rds_ file in a public facing Azure storage blob
* An _rds_ file is a compressed, serialized R object
* Save an object to _rds_ by using the `saveRDS` function; read an _rds_ object with the `readRDS` object

## Viewing Data
### tibble

* `dplyr` includes a wrapper called `tbl_df` that adds an additional class attribute onto `data.frames` that provides some better data manipulation aesthetics (there's now a dedicated package [`tibble`](www.github.com/hadley/tibble) for this wrapper and it's class)
* Most noticeable differential between `tbl_df` and `data.frame`s is the console output: `tbl_df`s will only print what the current R console window can display
* Can change the default setting for number of displayed columns by changing the options parameter: `options(dplyr.width = Inf)` 

```{r}
library(dplyr)
taxi_file <- file.path("/home/remoteuser/Data/manhattan_df.rds")
taxi_df  <- readRDS(file = taxi_file)
(taxi_df <- tbl_df(taxi_df))
```

# Filtering and Reordering Data

## Subsetting Data

* `dplyr` makes subsetting by rows very easy
* The `filter` verb takes conditions for filtering rows based on conditions
* **every** `dplyr` function uses a data.frame/tbl as it's first argument
* Additional conditions are passed as new arguments (no need to make an insanely complicated expression, split em up!)

## Filter

```{r}
filter(taxi_df,
       dropoff_dow %in% c("Fri", "Sat", "Sun"),
       tip_amount > 1)
```

## Exercise

Your turn: 

* How many observations started in Harlem?
  - pick both sides of Harlem, including east harlem
  - *hint*: it might be useful to use the `str_detect` function from `stringr`
* How many observations that started in Harlem ended in the Financial District?

## Solution

```{r}
library(stringr)
harlem_pickups <- filter(taxi_df, str_detect(pickup_nhood, "Harlem"))
harlem_pickups
# uncomment the line below (ctrl+shift+c) and filter harlem_pickups on Financial District
# how many rows?
# fidi <- filter(harlem_pickups, ...)
```

## Select a set of columns

* You can use the `select()` verb to specify which columns of a dataset you want
* This is similar to the `keep` option in SAS's data step.
* Use a colon `:` to select all the columns between two variables (inclusive)
* Use `contains` to take any columns containing a certain word/phrase/character

## Select Example

```{r}
select(taxi_df, pickup_nhood, dropoff_nhood, 
       fare_amount, dropoff_hour, trip_distance)
```

## Select: Other Options

starts_with(x, ignore.case = FALSE)

:    name starts with `x`

ends_with(x, ignore.case = FALSE)

:    name ends with `x`

matches(x, ignore.case = FALSE)

:    selects all variables whose name matches the regular expression `x`

num_range("V", 1:5, width = 1)

:    selects all variables (numerically) from `V1` to `V5`.

* You can also use a `-` to drop variables.

## Reordering Data

* You can reorder your dataset based on conditions using the `arrange()` verb
* Use the `desc` function to sort in descending order rather than ascending order (default)

## Arrange

```{r}
select(arrange(taxi_df, desc(fare_amount), pickup_nhood), 
       fare_amount, pickup_nhood)

head(select(arrange(taxi_df, desc(fare_amount), pickup_nhood), 
       fare_amount, pickup_nhood), 10)
```




## Exercise
Use `arrange()` to  sort on the basis of `tip_amount`, `dropoff_nhood`, and `pickup_dow`, with descending order for tip amount

## Summary

filter

:    Extract subsets of rows. See also `slice()`

select

:    Extract subsets of columns. See also `rename()`

arrange

:    Sort your data

# Data Aggregations and Transformations

## Transformations

* The `mutate()` verb can be used to make new columns

```{r}
taxi_df <- mutate(taxi_df, tip_pct = tip_amount/fare_amount)
select(taxi_df, tip_pct, fare_amount, tip_amount)
transmute(taxi_df, tip_pct = tip_amount/fare_amount)
```

## Summarise Data by Groups

* The `group_by` verb creates a grouping by a categorical variable
* Functions can be placed inside `summarise` to create summary functions

```{r}
grouped_taxi <- group_by(taxi_df, dropoff_nhood)
class(grouped_taxi)
grouped_taxi
```


```{r}
summarize(group_by(taxi_df, dropoff_nhood), 
          count = n(), ave_tip_pct = mean(tip_pct))
```

## Chaining/Piping

* A `dplyr` installation includes the `magrittr` package as a dependency 
* The `magrittr` package includes a pipe operator that allows you to pass the current dataset to another function
* This makes interpreting a nested sequence of operations much easier to understand

## Standard Code

* Code is executed inside-out.
* Let's arrange the above average tips in descending order, and only look at the locations that had at least 10 dropoffs and pickups.

```{r}
filter(arrange(summarise(group_by(taxi_df, pickup_nhood, dropoff_nhood), count = n(), ave_tip_pct = mean(tip_pct)), desc(ave_tip_pct)), count >= 10)
```

--- 


## Reformatted

```{r}
filter(
  arrange(
    summarise(
      group_by(taxi_df, 
               pickup_nhood, dropoff_nhood), 
      count = n(), 
      ave_tip_pct = mean(tip_pct)), 
    desc(ave_tip_pct)), 
  count >= 10)
```

## Magrittr

* Inspired by unix `|`, and F# forward pipe `|>`, `magrittr` introduces the funny character (`%>%`, the _then_ operator)
* `%>%` pipes the object on the left hand side to the first argument of the function on the right hand side
* Every function in `dplyr` has a slot for `data.frame/tbl` as it's first argument, so this works beautifully!


```{r}
taxi_df %>% 
  group_by(pickup_nhood, dropoff_nhood) %>% 
  summarize(count = n(),
            ave_tip_pct = mean(tip_pct)) %>% 
  arrange(desc(ave_tip_pct)) %>% 
  filter(count >= 10)
```

---

## Pipe and Plot

Piping is not limited to dplyr functions, can be used everywhere!

```{r, eval = FALSE, fig.height = 8, fig.width = 8}
library(ggplot2)
taxi_df %>% 
  group_by(dropoff_nhood, pickup_nhood) %>% 
  summarize(ave_tip = mean(tip_pct), 
            ave_dist = mean(trip_distance)) %>% 
  filter(ave_dist > 25, ave_tip > .5) %>% 
  ggplot(aes(x = pickup_nhood, y = dropoff_nhood)) + 
    geom_tile(aes(fill = ave_tip), colour = "white") + 
    theme_bw() + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = 'bottom') +
    scale_fill_gradient(low = "white", high = "steelblue")

```

---

## Piping to other arguments

* Although `dplyr` takes great care to make it particularly amenable to piping, other functions may not reserve the first argument to the object you are passing into it.
* You can use the special `.` placeholder to specify where the object should enter

```{r}
taxi_df %>% 
  group_by(dropoff_nhood, pickup_nhood) %>% 
  summarize(ave_tip = mean(tip_pct), 
            ave_dist = mean(trip_distance)) %>% 
  lm(ave_tip ~ ave_dist, data = .) -> taxi_model
summary(taxi_model)
```

## Exercise
  
Your turn: 

* Use the pipe operator to group by day of week and dropoff neighborhood
* Filter to Manhattan neighborhoods 
* Make tile plot with average fare amount in dollars as the fill



## Creating Complex Pipelines with do

* The `summarize` function is fun, can summarize many numeric/scalar quantities
* But what if you want multiple values/rows back, not just a scalar summary?
* Meet the `do` verb -- arbitrary `tbl` operations

---

```{r eval=FALSE}
taxi_df %>% group_by(dropoff_dow) %>%
  filter(!is.na(dropoff_nhood), !is.na(pickup_nhood)) %>%
  arrange(desc(tip_pct)) %>% 
  do(slice(., 1:2)) %>% 
  select(dropoff_dow, tip_amount, tip_pct, 
         fare_amount, dropoff_nhood, pickup_nhood)
```

## Estimating Multiple Models with do

* A common use of `do` is to calculate many different models by a grouping variable

```{r}
dow_lms <- taxi_df %>% sample_n(10^4) %>% 
  group_by(dropoff_dow) %>% 
  do(lm_tip = lm(tip_pct ~ pickup_nhood + passenger_count + pickup_hour,
     data = .))
```

---

```{r}
dow_lms
```


## Summary

mutate

:    Create transformations

summarise

:    Aggregate

group_by

:    Group your dataset by levels

do

:    Evaluate complex operations on a tbl

Chaining with the `%>%` operator can result in more readable code.
